---
title: "September Programming Project"
author: "Mattia Marcarini"
date: "`25-09-2025`"
output: html_document
---

# Introduction

In this analysis, two RNA-seq datasets (GSE244485 and GSE244486) are integrated and processed.  
The goal is to correct for batch effects, perform differential expression analysis, and compare coding and non-coding gene expression.

The main steps include:

1. **Gene Annotation** – preparation of genomic annotations from Ensembl GTF files.  
2. **PCA** – exploratory analysis to assess the presence of batch effects.  
3. **Batch Correction & Differential Expression** – batch effect correction with *ComBat-seq* and DE analysis with DESeq2.  
4. **Heatmap & Clustering** – visualization of the most variable genes after correction.  
5. **Coding vs Noncoding Genes** – comparison and visualization of top DE genes with violin plots.   


```{r step1, message=TRUE}
###############################################
# STEP 1: Gene Annotation
###############################################
# In this step, we download the GTF file from Ensembl,
# extract gene annotations, and create 3 objects:
# - gr_all: all genes
# - gr_coding: only protein-coding genes
# - gr_noncoding: only non-coding genes


library(GenomicRanges)
library(GenomeInfoDb)
library(S4Vectors)
library(IRanges)
library(rtracklayer)

# Creation of Data folder if it doesn't exist
if(!dir.exists("data")) dir.create("data")

# Creation of results folder if it doesn't exist
if(!dir.exists("results")) dir.create("results")

# Downloading GTF from Ensembl Release 114
gtf_file <- "data/Homo_sapiens.GRCh38.114.gtf.gz"
if(!file.exists(gtf_file)) {
message("⬇️ Downloading GTF from Ensembl...")
download.file(
url = "https://ftp.ensembl.org/pub/release-114/gtf/homo_sapiens/Homo_sapiens.GRCh38.114.gtf.gz
",
destfile = gtf_file
)
message("✅ Download completed.")
}

# Reading GTF in GRanges
gtf <- import(gtf_file)
genes <- gtf[gtf$type == "gene"]

# Keeping only genes with valid gene_id 
genes <- genes[!is.na(mcols(genes)$gene_id) & mcols(genes)$gene_id != ""]

# Renaming coding genes as ensemblID:geneSymbol
gene_biotype <- mcols(genes)$gene_biotype
gene_id     <- mcols(genes)$gene_id
gene_name   <- mcols(genes)$gene_name

# Updating gene_id column in GRanges
mcols(genes)$gene_id_updated <- ifelse(
  gene_biotype == "protein_coding",
  paste0(gene_id, ":", gene_name),  # coding: ensemblID:geneSymbol
  gene_id                           # noncoding: solo ensemblID
)

# Replace the gene_id column with the updated one

mcols(genes)$gene_id <- mcols(genes)$gene_id_updated
mcols(genes)$gene_id_updated <- NULL

# Saving GRanges complete
gr_all <- genes
save(gr_all, file = "data/gr_all.RData")

# Dividing into in coding and noncoding
gr_coding    <- gr_all[gene_biotype == "protein_coding"]
gr_noncoding <- gr_all[gene_biotype != "protein_coding"]

# Printing info
cat("✅ Total genes:", length(gr_all), "\n")
cat("🧬 Coding genes:", length(gr_coding), "\n")
cat("🌀 Noncoding genes:", length(gr_noncoding), "\n")

# Saving  subsets
save(gr_coding, file = "data/gr_coding.RData")
save(gr_noncoding, file = "data/gr_noncoding.RData")
cat("💾 Files saved in data/: gr_all.RData, gr_coding.RData, gr_noncoding.RData\n")
```

```{r step2, message=TRUE}
###############################################
# STEP 2: PCA to assess Batch Effect
###############################################
# PCA (Principal Component Analysis) allows to reduce dimensionality
# and observe if samples cluster by biological condition
# or by batch (indicating a batch effect).

# Load libraries

library(dplyr)
library(ggplot2)
library(Biobase)

# Paths to the TSV files in the data subfolder

file1 <- "data/GSE244486_raw_counts.tsv"
file2 <- "data/GSE244485_raw_counts.tsv"

# Read datasets
expr1 <- read.table(file1, header = TRUE, sep = "\t", row.names = 1)
expr2 <- read.table(file2, header = TRUE, sep = "\t", row.names = 1)

cat("Dimensions expr1:", dim(expr1), "\n")
cat("Dimensions expr2:", dim(expr2), "\n")

# Find common genes
common_genes <- intersect(rownames(expr1), rownames(expr2))
cat("Common genes:", length(common_genes), "\n")

if(length(common_genes) == 0){
  stop("❌ No gene in common between the two datasets. Check platform/annotation.")
}

# Keep only common genes
expr1_common <- expr1[common_genes, ]
expr2_common <- expr2[common_genes, ]

# Combine datasets
combined_expr <- cbind(expr1_common, expr2_common)

# Transpose for PCA (samples as rows)
combined_expr_t <- t(combined_expr)

# PCA
pca_res <- prcomp(combined_expr_t, scale. = TRUE)

# Create dataframe for ggplot
pca_df <- data.frame(
  PC1 = pca_res$x[,1],
  PC2 = pca_res$x[,2],
  Dataset = rep(c("GSE244486", "GSE244485"),
                times = c(ncol(expr1_common), ncol(expr2_common)))
)

# PCA plot
p <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Dataset)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "PCA of GEO datasets", x = "PC1", y = "PC2")

ggsave("results/PCA_plot.png", plot = p, width = 6, height = 5)  # save as PNG

# Calculate centroids of the two datasets
centroids <- pca_df %>%
  group_by(Dataset) %>%
  summarise(PC1_mean = mean(PC1), PC2_mean = mean(PC2))

# Euclidean distance between centroids
dist_centroids <- sqrt((centroids$PC1_mean[1] - centroids$PC1_mean[2])^2 +
                         (centroids$PC2_mean[1] - centroids$PC2_mean[2])^2)

cat("Distance between dataset centroids (batch effect proxy):", dist_centroids, "\n")

if(dist_centroids > 5){
  cat("⚠️ Likely relevant batch effect.\n")
} else {
  cat("✅ Minimal batch effect.\n")
}

print(p) # Show PCA inline in the report

```

```{r step3, message=TRUE}
###############################################
# STEP 3: Batch Correction and Differential Expression
###############################################
# Batch correction with ComBat-seq and DE analysis with DESeq2

# Step 3.1: If a batch effect is detected (prepare metadata)

library(dplyr)

# Input files
file1 <- "data/GSE244486_raw_counts.tsv"
file2 <- "data/GSE244485_raw_counts.tsv"

# Read the two datasets
expr1 <- read.table(file1, header = TRUE, sep = "\t", row.names = 1)
expr2 <- read.table(file2, header = TRUE, sep = "\t", row.names = 1)

# Rename columns with batch suffix
colnames(expr1) <- paste0(colnames(expr1), "_486")
colnames(expr2) <- paste0(colnames(expr2), "_485")

# Save renamed versions
write.table(expr1, "data/GSE244486_raw_counts_renamed.tsv",
            sep = "\t", quote = FALSE, col.names = NA)
write.table(expr2, "data/GSE244485_raw_counts_renamed.tsv",
            sep = "\t", quote = FALSE, col.names = NA)

# Build metadata
samples <- c(colnames(expr1), colnames(expr2))
conditions <- gsub("_.*", "", samples)  # take the part before "_"

metadata <- data.frame(
  sample = samples,
  condition = conditions
)

# Save metadata
write.table(metadata, "data/sample_metadata.tsv",
            sep = "\t", quote = FALSE, row.names = FALSE)

cat("✅ Created:\n - data/GSE244486_raw_counts_renamed.tsv\n - data/GSE244485_raw_counts_renamed.tsv\n - data/sample_metadata.tsv\n")

# Step 3.2: If a batch effect is detected 

library(data.table)
library(dplyr)
library(sva)
library(DESeq2)

# --- 1: Upload counts ---
counts_file_486 <- "data/GSE244486_raw_counts_renamed.tsv"
counts_file_485 <- "data/GSE244485_raw_counts_renamed.tsv"

counts_486 <- fread(counts_file_486, data.table = FALSE)
rownames(counts_486) <- counts_486[[1]]
counts_486 <- counts_486[,-1]

counts_485 <- fread(counts_file_485, data.table = FALSE)
rownames(counts_485) <- counts_485[[1]]
counts_485 <- counts_485[,-1]

counts_combined <- cbind(counts_486, counts_485)
colnames(counts_combined) <- make.unique(colnames(counts_combined))
cat("Dimensions of combined counts:", dim(counts_combined), "\n")

# --- 1b: Load metadata ---
metadata_file <- "data/sample_metadata.tsv"
metadata <- fread(metadata_file, data.table = FALSE)
rownames(metadata) <- metadata$sample
cat("Dimensions of metadata:", dim(metadata), "\n")

# Keep only common samples
common_samples <- intersect(colnames(counts_combined), rownames(metadata))
counts_combined <- counts_combined[, common_samples]
metadata <- metadata[common_samples, , drop=FALSE]
cat("Common samples:", length(common_samples), "\n")

# --- 2: Batch correction with ComBat-seq ---
metadata$batch <- ifelse(grepl("_486", rownames(metadata)), "486", "485")
combat_counts <- ComBat_seq(as.matrix(counts_combined), batch=metadata$batch, group=metadata$condition)
cat("Dimensions of combat_counts:", dim(combat_counts), "\n")

# --- 3: DESeq2 on COMBAT-corrected data ---
dds_combat <- DESeqDataSetFromMatrix(countData = round(combat_counts),
                                     colData = metadata,
                                     design = ~ condition)
dds_combat <- DESeq(dds_combat)

res_infected_combat <- results(dds_combat, contrast=c("condition","infected","mock"))
res_bystander_combat <- results(dds_combat, contrast=c("condition","bystander","mock"))

write.csv(as.data.frame(res_infected_combat), file="results/results_DE_infected_vs_mock_COMBAT.csv", row.names=TRUE)
write.csv(as.data.frame(res_bystander_combat), file="results/results_DE_bystander_vs_mock_COMBAT.csv", row.names=TRUE)
cat("DE results COMBAT saved as CSV\n")

# --- 4: DESeq2 on RAW data including batch as covariate ---
dds_raw <- DESeqDataSetFromMatrix(countData = round(counts_combined),
                                  colData = metadata,
                                  design = ~ batch + condition)
dds_raw <- DESeq(dds_raw)

res_infected_raw <- results(dds_raw, contrast=c("condition","infected","mock"))
res_bystander_raw <- results(dds_raw, contrast=c("condition","bystander","mock"))

write.csv(as.data.frame(res_infected_raw), file="results/results_DE_infected_vs_mock_RAW.csv", row.names=TRUE)
write.csv(as.data.frame(res_bystander_raw), file="results/results_DE_bystander_vs_mock_RAW.csv", row.names=TRUE)
cat("DE results RAW+batch saved as CSV\n")

# --- 5: Compare COMBAT vs RAW+batch ---
# Convert to data.frame and add gene names
df_combat_infected <- as.data.frame(res_infected_combat)
df_combat_infected$gene <- rownames(df_combat_infected)

df_raw_infected <- as.data.frame(res_infected_raw)
df_raw_infected$gene <- rownames(df_raw_infected)

compare_infected <- merge(df_combat_infected[, c("gene", "log2FoldChange", "padj")],
                          df_raw_infected[, c("gene", "log2FoldChange", "padj")],
                          by="gene",
                          suffixes=c("_combat", "_raw"))

write.csv(compare_infected, file="results/compare_DE_infected_vs_mock.csv", row.names=FALSE)
cat("Comparison DE infected vs mock COMBAT vs RAW saved\n")

df_combat_bystander <- as.data.frame(res_bystander_combat)
df_combat_bystander$gene <- rownames(df_combat_bystander)

df_raw_bystander <- as.data.frame(res_bystander_raw)
df_raw_bystander$gene <- rownames(df_raw_bystander)

compare_bystander <- merge(df_combat_bystander[, c("gene", "log2FoldChange", "padj")],
                           df_raw_bystander[, c("gene", "log2FoldChange", "padj")],
                           by="gene",
                           suffixes=c("_combat", "_raw"))

write.csv(compare_bystander, file="results/compare_DE_bystander_vs_mock.csv", row.names=FALSE)
cat("Comparison DE bystander vs mock COMBAT vs RAW saved\n")

# --- Save DESeq2 objects ---
save(dds_combat, file="results/dds_combat.RData")
save(dds_raw, file="results/dds_raw.RData")
cat("DESeq2 objects saved: dds_combat and dds_raw\n")

# Show a preview of results inline
knitr::kable(head(as.data.frame(res_infected_combat), 10), caption = "Top 10 DE genes (infected vs mock - COMBAT)")
knitr::kable(head(as.data.frame(res_bystander_combat), 10), caption = "Top 10 DE genes (bystander vs mock - COMBAT)")


```

```{r step4, message=TRUE}
###############################################
# STEP 4: Heatmap of the most variable genes
###############################################
# Visualization of the top 100 most variable genes

# Step 4: Heat Map and Clustering

library(DESeq2)
library(pheatmap)
library(matrixStats)
library(grid)
library(RColorBrewer)

# --- Load dds_combat ---
load("results/dds_combat.RData")  # contains dds_combat

# --- Variance stabilizing transformation ---
vsd <- vst(dds_combat, blind = FALSE)

# --- Select top 100 most variable genes ---
rv <- matrixStats::rowVars(assay(vsd), useNames = TRUE)
select <- order(rv, decreasing = TRUE)[1:100]

mat <- assay(vsd)[select, ]
mat <- mat - rowMeans(mat)

# --- Metadata for annotations ---
anno <- as.data.frame(colData(dds_combat)[, c("condition","batch")])

# --- More readable color palette ---
col_fun <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100)

# --- Create heatmap with more space for dendrograms ---
hm <- pheatmap(mat,
               annotation_col = anno,
               show_rownames = FALSE,
               clustering_distance_rows = "euclidean",
               clustering_distance_cols = "euclidean",
               clustering_method = "complete",
               color = col_fun,
               main = "Top 100 variable genes - COMBAT corrected",
               silent = TRUE,
               treeheight_row = 50,   # row dendrogram height
               treeheight_col = 50,   # column dendrogram height
               fontsize_row = 6,      # row text size
               fontsize_col = 8       # column text size
)

# --- Save to PDF and PNG ---
pdf("results/heatmap_top100.pdf", width=12, height=10)  # enlarged
grid::grid.newpage()
grid::grid.draw(hm$gtable)
dev.off()

png("results/heatmap_top100.png", width=1500, height=1200, res=150)
grid::grid.newpage()
grid::grid.draw(hm$gtable)
dev.off()

cat("✅ Heatmap saved in results/heatmap_top100.pdf\n")
# Show heatmap inline
pheatmap(mat,
annotation_col = anno,
show_rownames = FALSE,
clustering_distance_rows = "euclidean",
clustering_distance_cols = "euclidean",
clustering_method = "complete",
color = col_fun,
main = "Top 100 variable genes - COMBAT corrected")

```

```{r step5, message=TRUE}
###############################################
# STEP 5: Coding vs Noncoding Genes - Violin Plots
###############################################
# Selection of top 5 DE genes for each contrast
# Visualization of expression distribution with violin plots

library(DESeq2)
library(GenomicRanges)
library(BiocGenerics)
library(SummarizedExperiment)
library(dplyr)
library(tidyr)
library(ggplot2)

# ----- 1️⃣ Load DE results from CSV -----
cat("✅ Loading DE results...\n")
de_infected <- read.csv("results/results_DE_infected_vs_mock_COMBAT.csv", row.names = 1)
de_bystander <- read.csv("results/results_DE_bystander_vs_mock_COMBAT.csv", row.names = 1)
cat("✅ DE results loaded successfully\n")

# ----- 2️⃣ Load GenomicRanges -----
cat("✅ Loading GenomicRanges...\n")
load("data/gr_coding.RData")      # coding genes
load("data/gr_noncoding.RData")   # noncoding genes
cat("✅ GenomicRanges loaded successfully\n")

# ----- 3️⃣ Load dds_combat object -----
cat("✅ Loading dds_combat...\n")
load("results/dds_combat.RData")  # contains dds_combat
cat("✅ dds_combat loaded successfully\n")

# ----- 4️⃣ VST transformation -----
cat("✅ Calculating VST...\n")
vst_obj <- vst(dds_combat, blind = FALSE)
cat("✅ VST calculated\n")

# ----- 5️⃣ Extract top 5 significant DE genes -----
cat("✅ Cleaning gene IDs and selecting top 5...\n")
# Filter significant DE
de_genes_infected <- rownames(de_infected)[de_infected$padj < 0.05]
de_genes_bystander <- rownames(de_bystander)[de_bystander$padj < 0.05]

# Intersect with genes present in VST
de_genes_infected <- intersect(de_genes_infected, rownames(vst_obj))
de_genes_bystander <- intersect(de_genes_bystander, rownames(vst_obj))

# Order by absolute log2FoldChange and take top 5
top5_infected <- head(de_genes_infected[order(abs(de_infected[de_genes_infected, "log2FoldChange"]), decreasing = TRUE)], 5)
top5_bystander <- head(de_genes_bystander[order(abs(de_bystander[de_genes_bystander, "log2FoldChange"]), decreasing = TRUE)], 5)
cat("✅ Top 5 genes selected\n")

# ----- 6️⃣ Function for violin plot -----
plot_violin <- function(genes, vst_obj, group_col = "condition", title = "Violin Plot") {
  mat <- assay(vst_obj)[genes, , drop = FALSE]
  df <- as.data.frame(t(mat))
  df$sample <- rownames(df)
  
  df_long <- pivot_longer(df, cols = -sample, names_to = "gene", values_to = "expression")
  df_long$group <- colData(vst_obj)[df_long$sample, group_col]
  
  p <- ggplot(df_long, aes(x = gene, y = expression, fill = group)) +
    geom_violin(trim = FALSE) +
    geom_boxplot(width = 0.1, fill = "white") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = title, y = "VST expression", x = "Gene")
  
  return(p)
}

# ----- 7️⃣ Generate plots for top 5 -----
cat("✅ Generating top 5 violin plots...\n")
p_infected <- plot_violin(top5_infected, vst_obj, group_col = "condition", title = "Top 5 Infected vs Mock")
p_bystander <- plot_violin(top5_bystander, vst_obj, group_col = "condition", title = "Top 5 Bystander vs Mock")

# Save
ggsave("results/violin_top5_infected_vs_mock.png", p_infected, width = 10, height = 6)
ggsave("results/violin_top5_bystander_vs_mock.png", p_bystander, width = 10, height = 6)
cat("✅ Top 5 violin plots saved in results/\n")

# Show violin plots inline
print(p_infected)
print(p_bystander)



```

# Reproducibility in Docker

This section describes how to reproduce the analysis performed in this report inside Docker.

Start your Docker

Using command prompt navigate to the project folder (which should contain `SeptemberProject.rmd`, `data/`, and `results/`) with:
cd /path/to/project

Build the docker image using the Dockerfile in the project from command prompt with:

docker build -t rna-seq-pipeline .

Start the Docker container with:
docker run -it --rm -v %cd%:/home/project rna-seq-pipeline

(On Linux/macOS, replace `%cd%` with `$(pwd)`)

Inside the container, start R with:
R

Then render the report with:
library(rmarkdown)
render("SeptemberProject.rmd")

The file `SeptemberProject.html` will be generated in the project folder on the host computer.

Alternatively, without entering R interactively, you can run directly:
docker run -it --rm -v %cd%:/home/project rna-seq-pipeline Rscript -e "rmarkdown::render('SeptemberProject.rmd')"

The plots and tables will be visible in the `SeptemberProject.html` file, and the files generated in the `results/` folder will also be accessible on the host computer.




## 🔧 Issues & Workarounds

1. **Missing `data/` folder**  
   - *Issue*: On the first run, some scripts could not find the `data/` folder.  
   - *Workaround*: Added a check  
     ```r
     if(!dir.exists("data")) dir.create("data")
     ```  
     to automatically create the folder.  

2. **Missing or not downloaded GTF file**  
   - *Issue*: The Ensembl GTF file was not available locally.  
   - *Workaround*: Conditional download  
     ```r
     if(!file.exists(gtf_file)) download.file(url, gtf_file)
     ```  
     to fetch the correct Ensembl release.  

3. **Incomplete gene annotation (`NA` in `gene_id`)**  
   - *Issue*: Some records in the GTF contained missing or empty `gene_id`.  
   - *Workaround*: Filtered out with  
     ```r
     genes <- genes[!is.na(mcols(genes)$gene_id) & mcols(genes)$gene_id != ""]
     ```  

4. **Duplicate columns in raw counts from GEO datasets**  
   - *Issue*: After `cbind` of GSE244486 and GSE244485, some sample names were duplicated.  
   - *Workaround*: Applied  
     ```r
     colnames(counts_combined) <- make.unique(colnames(counts_combined))
     ```  
     to ensure unique column names.  

6. **Sample alignment between counts and metadata**  
   - *Issue*: Mismatch between `colnames(counts_combined)` and `rownames(metadata)`.  
   - *Workaround*: Kept only common samples with  
     ```r
     common_samples <- intersect(colnames(counts_combined), rownames(metadata))
     counts_combined <- counts_combined[, common_samples]
     metadata <- metadata[common_samples, , drop=FALSE]
     ```  

7. **Plots with unreadable labels**  
   - *Issue*: Dendrograms and labels were too small in the heatmap.  
   - *Workaround*: Adjusted parameters  
     ```r
     treeheight_row = 50
     treeheight_col = 50
     fontsize_row = 6
     fontsize_col = 8
     ```  
     and exported PDF/PNG in high resolution.  

8. **Selection of top DE genes for violin plots**  
   - *Issue*: Some significant genes were missing in the `vst` object.  
   - *Workaround*: Intersected `rownames(vst_obj)` with DE genes filtered by `padj < 0.05`.  


